Seagrass mapping and monitoring \\ along the coasts of Crete, Greece

### 2011-02-05 MScDiplom_Lemenkova.tex
changed page layout to ISO-216 PA5 (140*210mm)

\documentclass[11pt, a4paper]{article}
\usepackage{authblk}
%\usepackage{authorindex}
\usepackage{color}
\usepackage[stable]{footmisc }
\usepackage{setspace}
\usepackage[printonlyused,withpage]{acronym}
\usepackage[a4paper, left=30mm,right=25mm,top=30mm,bottom=35mm]{geometry}
%\usepackage{glossary}
\usepackage{graphics}
\usepackage{graphicx}
\usepackage[numbers]{natbib}
\usepackage{calc}
\usepackage{amsmath}
\usepackage{longtable}
\usepackage[svgnames,table]{xcolor}
\usepackage[subfigure]{tocloft}
%\usepackage[table]{xcolor}
\usepackage{paralist}
%\usepackage{makeidx}\makeindex
%\usepackage[makeindex,protected]{splitidx}
%\usepackage{index}
\usepackage{imakeidx}
\makeindex[name=concepts,title=Index of Concepts,intoc]
\makeindex[name=people,title=Index of People,intoc,columns=3,columnseprule]
\usepackage{fancyhdr}
\pagestyle{fancy}
\usepackage{fancyheadings}
\topmargin = 1pt
\headheight=10mm
\headsep=7pt
%\ensuremath
\usepackage{lastpage}
\usepackage{float}
\usepackage{subfig}
\usepackage{caption}
\usepackage{chngcntr}
\counterwithin{figure}{section}
\counterwithin{table}{section}
\usepackage{appendix}
\captionsetup{margin=10pt,font=small,labelfont=bf}
\usepackage{wrapfig}
\definecolor{darkblue}{rgb}{0.0,0.0,0.3}
\setcounter{tocdepth}{4}
\parindent 0pt
%\floatstyle{boxed} 
\restylefloat{figure}
\usepackage[a4paper=true, ]{hyperref}
\hypersetup{pdftitle={Seagrass mapping and monitoring along the coasts of Crete, Greece}, pdfauthor={Lemenkova Polina}, pdfsubject={MScThesis}, pdfcreator={Lemenkova Polina}, pdfproducer={Lemenkova Polina}, colorlinks=true,linkcolor=Navy, citecolor=Crimson, filecolor=magenta, urlcolor=blue}

@@

\textbf{Course Duration:} \hfill{September 2009 – March 2011}\\

@@

\begin{itemize}
	\item[*] \href{http://www.southampton.ac.uk/}{University of Southampton}, \href{http://www.southampton.ac.uk/geography/}{School of Geography} (UK)\\
	\item[*] \href{http://www.lunduniversity.lu.se/}{Lund University}, \href{http://www.giscentrum.lu.se/english/index.htm}{GIS Centre} (Sweden)\\
	\item[*] \href{http://www.uw.edu.pl/en/}{University of Warsaw} (Poland)\\
	\item[*] \href{http://www.utwente.nl/en}{University of Twente}, \href{http://www.itc.nl/}{Faculty of Geo-Information Science and Earth Observation (ITC)} (The Netherlands)
\end{itemize}

@@

This research was carried out at the University of Twente, \href{http://www.itc.nl/}{Faculty of Geo-Information Science and Earth Observation (ITC)}, Enschede, the Netherlands, under the financial support of the EU \href{http://ec.europa.eu/education/external-relation-programmes/doc72_en.htm}{Erasmus Mundus Scholarship}. 
\vspace{1mm}\\
I thank my scientific supervisors, Dr. \textit{Albertus G.Toxopeus}\index[people]{Toxopeus, A.G.}  and Mr. \textit{Valentijn Venus}\index[people]{Venus, V.} from the ITC, Department of Natural Resources, for our successful scientific collaboration, help and discussions of my research work. \\
I also express them my most sincere appreciations for our fieldwork on Crete. \\ Besides scientific routine, we had an unforgettable time spent together. \\ Personally, I would like to mention that I have visited Greece the first time in my life, and I will always remember my Cretan period-2010 and Ligaria beach.\vspace{1mm}\\
I extend my gratitude to the staff of the Museum of Natural History, the University of Crete, especially to Dr.\textit{Petros Limberakis}\index[people]{Limberakis, Petros} and his colleagues, for their assistance during our fieldwork, providing us with all necessary equipment (e.g. SCUBA gear diving, instruments for underwater measurements, etc). I also appreciate their very warm, personal attitude and interest towards our research. 
\vspace{1mm}\\
My sincere gratitude goes again to Mr.\textit{Valentijn Venus}\index[people]{Venus, V.} for his generous providing me with a lot of spatial data   relevant to my research (e.g. such as digital vector layers of Crete island, raster georeferenced maps for the whole island for my GIS project, Excel spreadsheets, etc). Without these vital data the implementation of this project would have hardly been possible. I am also thankful to him for the technical help in Google Earth imagery capturing, using GDAL technologies and assistance during the fieldwork (videometric underwater footage, consultancies in seagrass observation and monitoring, driving boat, etc). \\ My special thank to his personal, friendly support in various situations on Crete.\vspace{1mm}\\
I am thankful to Dr.rer.nat.\textit{Peter Gege}\index[people]{Gege P.}, the software developer from the Earth Observation Centre of the German Space Agency, for his generous consultancies concerning the technical question about WASI that arose during my work. I am thankful for all his explanations and articles given to me via e-mail \vspace{1mm}\\
My special thanks are addressed to my elder brother, Mr.\textit{Peter Lemenkov}\index[people]{Lemenkov, P.}, the software developer from the ipPort, for his priceless numerous advices and constant professional support throughout my career. He encouraged me to use various very useful software and tools, always gave me necessary recommendations and precious help, thus highly contributing to my professional growth. \\ My special thank for his help with Python scripting. \\
I am also heartfelt thankful to him for his personal friendship and support. 
\vspace{1mm}\\
I am grateful to Mr.\textit{Willem Nieuwenhuis}\index[people]{Nieuwenhuis, W.}, the software developer from the ITC, Department of Natural Resources, for shared technology of Google Earth aerial imagery grabbing and for his consultancy with ASTER imagery. 
\vspace{1mm}\\
I thank Dr. Ir. \textit{Cornelis A.J.M. de Bie}\index[people]{de Bie C.A.J.M.} from the Department of Natural Resource, ITC, for his patient sharing of iPAQ knowledge during my Polish fieldwork in summer 2010. I used this skills on Crete. 
\vspace{1mm}\\
Many thanks for Dr. \textit{Dingtian Yang}\index[people]{Yang D.} from the South China Sea Institute of Oceanology, our colleague in problem of seagrass monitoring and research, for his handing over the data of spectral reflectance of Thalassia seagrass, which enabled to compare spectra of various seagrass types.

@@

	%\acro{GRETL}{GNU Regression, Econometrics and Time-series Library}

@@

\begin{wrapfigure}{r}{0.4\textwidth}
	\centering
	\includegraphics[scale=0.30]{Fig-1-2.jpg}
	\caption{Distribution of seagrasses in the world.\\ Source:\cite{Green03}\label{Green03} 
	\label{fig:2}} 
\end{wrapfigure}

@@

\begin{wrapfigure}{r}{0.4\textwidth}
	\centering
	\includegraphics[scale=0.25]{Fig-1-3.jpg}
	\caption{Morphology of different types of seagrasses. Source: \url{http://www.ozcoasts.org.au}}
	\label{fig:3}
\end{wrapfigure}

@@

\subsubsection{Environmental vulnerability of the seagrasses}
Meadows of \textit{P. oceanica} are subjected to the human activities\index[concepts]{Seagrasses!ecological vulnerability}, as they occur in coastal areas, where
they can be affected both directly \cite{Meinesz91}\label{Meinesz91} or indirectly, through the impact on the
quality of waters and sediments \cite{Duarte02}\label{Duarte02}. As \textit{P.oceanica} is a long-living plant with a slow
growth rate, the anthropogenic modifications of the coastal zone, happening more rapidly than the
capacity of the plant to adapt to these changes, reduce its distribution area \cite{Micheli05}\label{Micheli05}.

@@

One of the main drivers of seagrass decline is, for instance, the location of the fish farming near the
seagrass meadows. The negative effects of the sedimentation of waste particles in the farm vicinities
on \textit{P.oceanica} meadows are diverse and complex, and may cause benthic deterioration, accumulation
of organic matter and seagrass decline \cite{Holmer08}\label{Holmer08}. Seagrasses are subject to anthropogenic
nutrient (N and P) loading, which may occasionally casue morphological (e.g. leaf length) and
physiological (e.g. chlorophyll and nitrogen content of the leaves) responses towards changed
environmental conditions \cite{Leoni06,Leoni07}\label{Leoni06} \label{Leoni07}. 
The detailed research of the fish
farm-induced decline of the seagrass meadows \cite{Diaz-Almela06}\label{Diaz-Almela06}reports the relationships of
fish farm organic and nutrient content in the sediments with dynamics of the key seagrass species (\textit{P.
oceanica}) in the Mediterranean Sea. Nowadays \textit{P.oceanica} is in the alarming state of regression due
to the deterioration of the environment in the Mediterranean Sea \cite{Ribed02}\label{Ribed02}. Due to these reasons,
\textit{P.oceanica} is a protected species since 1988 in some European countries \cite{Francour99}\label{Francour99}, and
its presence serves as an indicator of a stable healthy environment. Among other negative factors,
affecting both growth and status of the seagrasses the environmental contaminants can be mentioned,
e.g. thermal, sewage, dredging and chemical pollution\index[concepts]{Seagrasses!anthropogenic impacts on}  as well as any other kind of maritime works,
e.g. trawling and anchoring of boats \cite{Ribed02}\label{Ribed02}. Other human activities that cause degrading of the
seagrass are recreational boating, commercial overexploitation of coastal resources, eutrophication
\cite{McKenzie06}\label{McKenzie06}.\\
Besides anthropogenic factors, various biochemical, climatic and environmental processes\index[concepts]{Seagrasses!natural impacts on} can cause
negative influence on seagrass distribution. Seagrass is exposed to threats from the global climate and
environmental change, i.e. increases in sea surface temperature; sea level rise; increased frequency
and intensity of storms and waves; local decrease of water quality, increased sedimentation,
contamination and nitrification; desiccation; salinity fluctuations; nutrient changes; suspended
sediments \cite{Blake00}\label{Blake00}. These stress-drivers can alone result in large-scale seagrass
degradation, but often seagrass undergo simultaneous affects from several of these factors together.
It naturally increases the environmental pressing  to drastic loss of very large areas of
seagrass globally \cite{Orth06}\label{Orth06}.
In tropical areas, where most of seagrasses are located (Fig.\ref{Fig.4}), seagrasses are subjectto catastrophic
extinction and loss, due to the cyclones, typhoons, storms, regular floods and increased rainfalls. Recovery from such
events can take up to several years and often it is only possible by means of the seed reserves from the local
environmental surveys \cite{McKenzie07}\label{McKenzie07}. Other threats for seagrasses in tropical areas are 
increased nutrient availability in the coastal zones, increased
eutrophication and invasive macroalgae. These processes have strong affect on the status of the
seagrass meadows, and often lead to their complete disappearance \cite{Holmer09}\label{Holmer09}.\\
Other environmental threat for the seagrasses arises as a result of the environmental struggle and
competition for existence among species. Thus, meadows of P.oceanica in the Mediterranean Sea are
presently facing invasion by alien algal species, particularly in areas where \textit{P.oceanica} is already
degrading, stressed, have gaps and patchy structure in meadows and show other signs of regression
\cite{Montefalcone10}\label{Montefalcone10}.
Seagrasses are vulnerable fragile species, important for the marine coastal ecosystems, especially for
the protection of the beach structure. However, the facts about seagrass global degrading sound
worrying: about 54 \% of the total seagrass meadows have lost any part of their area; the areas, where
the seagrass ecosystems are degrading or lost, are not located in a specific area or continent, but
registered globally; since 1980s global losses of the seagrasses on our planet is equal to two football
fields per hour \cite{Mellors09b}\label{Mellors09b}.

@@

\subsection{Research problem}\index[concepts]{Research}
Monitoring\index[concepts]{Research!problem} of the marine benthic ecosystems of seagrasses is essential for the environmental
assessment of the coastal zones. It increases our knowledge of the seagrass ecology, highlights 
threats to the seagrass and preventing them from possible losses and degrading and improves
techniques and methods of the underwater-based observations. Mapping the seagrass contributes to
the evaluating of the seagrass current distribution, analysis of its dynamics and changes over time, as
well as estimations of the degrading of seagrass meadows for the purpose of the coastal management.\\
Precise, correct and up-to-date information about the distribution of \textit{P.oceanica} is necessary for the
sustainable conservation of the marine environment and ecosystems in Mediterranean area, being an
important contribution to the environmental coastal zone management \cite{Pergent-Martini06}\label{Pergent-Martini06}.
However, mapping the seagrass has limitations due to the specific location and characteristics of the
research object. The remote sensing techniques have traditionally been widely used for the seagrass
monitoring.

\begin{wrapfigure}{r}{0.4\textwidth}
	\centering
	\includegraphics[scale=0.25]{Fig-1-9.jpg}
	\caption{Spectra of the seagrass on different depths (0 - 15 m). Source \cite{Edwards00}\label{Edwards00}}
	\label{fig:9}
\end{wrapfigure}

@@

Universal, international, standardized methods for seagrass directly for seagrasses
as such still should be developed. Various case studies have been performed, yet their mostly report
methods adjusted for particular areas, without evaluating standard general algorithms that could be
extrapolated towards other regions. \\ Application of the remote sensing towards seagrass mapping is
generally based on the assumption that various types of the seafloor bottom have different
characteristics of the reflectivity, which is visually expressed in distinct colors of the objects. In its
turn, reflectivity of the sediments is affected by the water optical properties and content. For
example, Stephens \cite{Stephens03}\label{Stephens03} prove that microalgae biomass and community structure affect
hyperspectral reflectance of sediments, which enable to estimate total microalgae biomass from
measurements of hyperspectral reflectance. 
Spectral measurements\index[concepts]{Remote sensing!spectral measurements} of the target objects are made
by means of the radiometers, which receive and register the amounts of energy (radiance and
irradiance) from the objects. Measuring optical properties of of the seawater allows to calculate spectra of the
objects and to discriminate them on the aerial and satellite images. Thus, various scientists report
success in spectral discrimination of submerged vegetation and other seafloor cover types on imagery
using hyperspectral optical properties of the sea water for the assessment of benthic habitats \cite{Lewis01,Louchard03,Dehouck08,Werdell03}\label{Lewis01}\label{Louchard03}\label{Dehouck08}\label{Werdell03}.
Studies of spectral reflectance\index[concepts]{Seawater optical properties!spectral reflectance} of the different seagrass species comparing to the spectra of sand and
other seafloor cover types \cite{Vahtmae06}\label{Vahtmae06} prove that spectra of green, brown and red benthic
macroalgae differ from each other, as well as from sand and deep water reflectance spectra. These
differences are well detectable by the means of the remote sensing research methods. Comparing to
the terrestrial plants, aquatic vegetation, \textit{inter alia} seagrass, cannot be detected using red edge of the
spectrum, as these wavelengths are significantly absorbed by water \cite{Kirk94}\label{Kirk94}, as well as by
scattering and absorption by phytoplankton. Some authors \cite{Dierssen03}\label{Dierssen03} also use
spectrally based radiative transfer approach to quantitatively estimate shallow-water bathymetry and
leaf area index (\ac{LAI}) of the seagrass. The spectral reflectance in general is the
result of the spectral absorption in different bands, typical for each target object.

@@

Spectral reflectance\index[concepts]{Seawater optical properties!spectral reflectance} of the seagrass  (Fig.\ref{fig:13}) is largely influenced by the water
depth where it is located, and is generally decreasing in values by increasing depths.
The most important spectral diapason\index[concepts]{Seawater optical properties!spectral reflectance!spectral diapason} for marine mapping of submerged vegetation
and, particularly, for seagrass, is 350-800 nm \cite{Barille09}\label{Barille09}.
Using airborne imagery for retrospective data (before 1970s) together with the most
recent imagery allows to detect changes in seagrass distribution on over different years
and to analyse dynamics of the seagrass distribution \cite{Ball09}\label{Ball09}. \\
Another important advantage of the application of the remote sensing data for mapping of submerged aquatic vegetation has
commercial nature: using remote sensing data and methods enables more low-cost and up-to-date
seagrass mapping \cite{Mumby99}\label{Mumby99}, and is especially useful for the areas where the fieldwork data
capturing is unavailable.
Seagrass meadows may reach spatial scales from several up to hundreds of metres, therefore they are
susceptible by the means of satellite imagery from remote sensors, both with moderate resolution
(e.g., \href{http://landsat.gsfc.nasa.gov/}{Landsat MSS, Landsat TM, Landsat ETM+}, \href{http://envisat.esa.int/instruments/meris/}{MERIS} from \ac{ESA}, \href{http://asterweb.jpl.nasa.gov/}{ASTER}, \href{http://modis.gsfc.nasa.gov/}{MODIS}) and high resolution as
well (e.g., \href{http://www.satimagingcorp.com/gallery-ikonos.html}{IKONOS}, \href{http://www.digitalglobe.com/index.php/85/QuickBird}{Quickbird}, \href{http://www.spotimage.fr/}{SPOT}, \href{http://www.itres.com/products/imagers/casi550}{CASI}). The possibility of their application towards seagrass mapping varies and is limited by the technical characteristics (Table \ref{tab:1}) and resolution of these sensors  (Fig.\ref{fig:2.3}).
In the next paragraphs (\nameref{sec:2.4.1} and \nameref{sec:2.4.2}) we will briefly discuss research experience of the application of satellite imagery towards seagrass mapping, its advantages and limitations, reported by various researchers.

\subsubsection[Review of multispectral imagery...]{Review of multispectral imagery used for seagrass mapping}\label{sec:2.4.1}\index[concepts]{Remote sensing!multispectral imagery} 
Seagrass mapping using remotely sensed data from multispectral sensors is based on the classification
and discrimination of the seafloor types using their spectral characteristics in different wavebands.
Perhaps, the most known imagery, widely used in the remote sensing mapping, is received from the
group of Landsat sensors, known for its historical, pioneer role in the satellite industry.
Being the longest running satellite system, lunched in 1972, Landsat is the only source of archival
data going back to 1984 at a sufficient spatial resolution \cite{Dekker05b}\label{Dekker05b}, which makes its data
desirable for historical mapping or environmental analysis of change detection of the seagrass
landscapes.

\begin{wrapfigure}{r}{0.5\textwidth}
	\centering
	\includegraphics[scale=0.35]{Fig-13.jpg}
 	\caption{Band coverage of ASTER and Landsat channels on the e/m spectrum. Source:\cite{Kalinowski04}\label{Kalinowski04}}
	\label{fig:2.4}
\end{wrapfigure}

>>>>

\documentclass[11pt]{article}
\special{papersize=140mm,210mm}
\usepackage{authblk}
%\usepackage{authorindex}
\usepackage{color}
\usepackage[stable]{footmisc }
\usepackage{setspace}
\usepackage[printonlyused,withpage]{acronym}
\usepackage[left=30mm,right=20mm,top=30mm,bottom=35mm]{geometry}
%\usepackage{glossary}
\usepackage{graphics}
\usepackage{graphicx}
\usepackage[numbers]{natbib}
\usepackage{calc}
\usepackage{amsmath}
\usepackage{longtable}
\usepackage[svgnames,table]{xcolor}
\usepackage[subfigure]{tocloft}
%\usepackage[table]{xcolor}
\usepackage{paralist}
%\usepackage{makeidx}\makeindex
%\usepackage[makeindex,protected]{splitidx}
%\usepackage{index}
\usepackage{imakeidx}
\makeindex[name=concepts,title=Index of Concepts,intoc]
\makeindex[name=people,title=Index of People,intoc,columns=3,columnseprule]
\usepackage{fancyhdr}
\pagestyle{fancy}
\usepackage{fancyheadings}
\topmargin = 1pt
\headheight=10mm
\headsep=7pt
%\ensuremath
\usepackage{lastpage}
\usepackage{float}
\usepackage{subfig}
\usepackage{caption}
\usepackage{chngcntr}
\counterwithin{figure}{section}
\counterwithin{table}{section}
\usepackage{appendix}
\captionsetup{margin=10pt,font=small,labelfont=bf}
\usepackage{wrapfig}
\definecolor{darkblue}{rgb}{0.0,0.0,0.3}
\setcounter{tocdepth}{4}
\parindent 0pt
%\floatstyle{boxed} 
\restylefloat{figure}
\usepackage{hyperref}
\hypersetup{pdftitle={Seagrass mapping and monitoring along the coasts of Crete, Greece}, pdfauthor={Lemenkova Polina}, pdfsubject={MScThesis}, pdfcreator={Lemenkova Polina}, pdfproducer={Lemenkova Polina}, colorlinks=true,linkcolor=Navy, citecolor=Crimson, filecolor=magenta, urlcolor=blue}

@@

\textbf{Course Duration:} \hfill{September 2009 $-$ March 2011}\\

@@

\begin{itemize}
	\item[*] \href{http://www.southampton.ac.uk/}{University of Southampton}, \href{http://www.southampton.ac.uk/geography/}{School of Geography} (UK)\\
	\item[*] \href{http://www.lunduniversity.lu.se/}{Lund University}, \href{http://www.giscentrum.lu.se/english/index.htm}{GIS Centre} (Sweden)\\
	\item[*] \href{http://www.uw.edu.pl/en/}{University of Warsaw} (Poland)\\
	\item[*] \href{http://www.utwente.nl/en}{University of Twente}, \href{http://www.itc.nl/}{Faculty of Geo-Information Science and Earth Observation (ITC)} \\ (The Netherlands)
\end{itemize}

@@

This research was carried out at the University of Twente, \href{http://www.itc.nl/}{Faculty of Geo-Information Science and Earth Observation (ITC)}, Enschede, the Netherlands, under the financial support of the EU \href{http://ec.europa.eu/education/external-relation-programmes/doc72_en.htm}{Erasmus Mundus Scholarship}. 
\vspace{1mm}\\
I thank my scientific supervisors, Dr. \textit{Albertus G.Toxopeus}\index[people]{Toxopeus, A.G.}  and Mr. \textit{Valentijn Venus}\index[people]{Venus, V.} from the ITC, Department of Natural Resources, for our successful scientific collaboration, help and discussions of my research work. 
I also express them my most sincere appreciations for our fieldwork on Crete. \\ Besides scientific routine, we had an unforgettable time spent together. \\ Personally, I would like to mention that I have visited Greece the first time in my life, and I will always remember my Cretan period-2010 and Ligaria beach.\vspace{1mm}\\
I extend my gratitude to the staff of the Museum of Natural History, the University of Crete, especially to Dr.\textit{Petros Limberakis}\index[people]{Limberakis, Petros} and his colleagues, for their assistance during our fieldwork, providing us with all necessary equipment (e.g. SCUBA gear diving, instruments for underwater measurements, etc). I also appreciate their very warm, personal attitude and interest towards our research. 
\vspace{1mm}\\
My sincere gratitude goes again to Mr.\textit{Valentijn Venus}\index[people]{Venus, V.} for his generous providing me with a lot of spatial data   relevant to my research (e.g. such as digital vector layers of Crete island, raster georeferenced maps for the whole island for my GIS project, Excel spreadsheets, etc). Without these vital data the implementation of this project would have hardly been possible. I am also thankful to him for the technical help in Google Earth imagery capturing, using GDAL technologies and assistance during the fieldwork (videometric underwater footage, consultancies in seagrass observation and monitoring, driving boat, etc). \\ My special thank to his personal, friendly support in various situations on Crete.\vspace{1mm}\\
I am thankful to Dr.rer.nat.\textit{Peter Gege}\index[people]{Gege P.}, the software developer from the Earth Observation Centre of the German Space Agency, for his generous consultancies concerning the technical question about WASI that arose during my work. I am thankful for all his explanations and articles given to me via e-mail \vspace{1mm}\\
My special thanks are addressed to my elder brother, Mr.\textit{Peter Lemenkov}\index[people]{Lemenkov, P.}, the software developer from the ipPort, for his priceless numerous advices and constant professional support throughout my career. He encouraged me to use various very useful software and tools, always gave me necessary recommendations and precious help, thus highly contributing to my professional growth. \\ My special thank for his help with Python scripting. \\
I am also heartfelt thankful to him for his personal friendship and support. 
\vspace{1mm}\\
I am grateful to Mr.\textit{Willem Nieuwenhuis}\index[people]{Nieuwenhuis, W.}, the software developer from the ITC, Department of Natural Resources, for shared technology of Google Earth aerial imagery grabbing and for his consultancy with ASTER imagery. 
\vspace{1mm}\\
I thank Dr. Ir. \textit{Cornelis A.J.M. de Bie}\index[people]{de Bie C.A.J.M.} from the Department of Natural Resource, ITC, for his patient sharing of iPAQ knowledge during my Polish fieldwork in summer 2010. I used this skills on Crete. 
\vspace{1mm}\\
Many thanks for Dr. \textit{Dingtian Yang}\index[people]{Yang D.} from the South China Sea Institute of Oceanology, our colleague in problem of seagrass monitoring and research, for his handing over the data of spectral reflectance of Thalassia seagrass, which enabled to compare spectra of various seagrass types.

@@

	\acro{GRETL}{\href{http://gretl.sourceforge.net/}{GNU Regression, Econometrics and Time-series Library}}

@@

\begin{wrapfigure}{r}{0.45\textwidth}
	\centering
	\includegraphics[scale=0.40]{Fig-1-2.jpg}
	\caption{Distribution of seagrasses in the world. Source:\cite{Green03}\label{Green03} 
	\label{fig:2}} 
\end{wrapfigure}

@@

\begin{wrapfigure}{r}{0.35\textwidth}
	\centering
	\includegraphics[scale=0.3]{Fig-1-3.jpg}
	\caption{Morphology of different types of seagrasses. Source: \url{http://www.ozcoasts.org.au}}
	\label{fig:3}
\end{wrapfigure}

@@

\subsubsection{Environmental vulnerability of the seagrasses}
Meadows of \textit{P. oceanica} are subjected to the human activities\index[concepts]{Seagrasses!ecological vulnerability}, as they occur in coastal areas, where
they can be affected both directly \cite{Meinesz91}\label{Meinesz91} or indirectly, through the impact on the
quality of waters and sediments \cite{Duarte02}\label{Duarte02}. As \textit{P.oceanica} is a long-living plant with a slow
growth rate, the anthropogenic modifications of the coastal zone, happening more rapidly than the
capacity of the plant to adapt to these changes, reduce its distribution area \cite{Micheli05}\label{Micheli05}. 
One of the main drivers of seagrass decline is, for instance, the location of the fish farming near the
seagrass meadows. The negative effects of the sedimentation of waste particles in the farm vicinities
on \textit{P.oceanica} meadows are diverse and complex, and may cause benthic deterioration, accumulation
of organic matter and seagrass decline \cite{Holmer08}\label{Holmer08}. Seagrasses are subject to anthropogenic
nutrient (N and P) loading, which may occasionally casue morphological (e.g. leaf length) and
physiological (e.g. chlorophyll and nitrogen content of the leaves) responses towards changed
environmental conditions \cite{Leoni06,Leoni07}\label{Leoni06} \label{Leoni07}. 
The detailed research of the fish
farm-induced decline of the seagrass meadows \cite{Diaz-Almela06}\label{Diaz-Almela06}reports the relationships of
fish farm organic and nutrient content in the sediments with dynamics of the key seagrass species (\textit{P.
oceanica}) in the Mediterranean Sea. Nowadays \textit{P.oceanica} is in the alarming state of regression due
to the deterioration of the environment in the Mediterranean Sea \cite{Ribed02}\label{Ribed02}. Due to these reasons,
\textit{P.oceanica} is a protected species since 1988 in some European countries \cite{Francour99}\label{Francour99}, and
its presence serves as an indicator of a stable healthy environment. Among other negative factors,
affecting both growth and status of the seagrasses the environmental contaminants can be mentioned,
e.g. thermal, sewage, dredging and chemical pollution\index[concepts]{Seagrasses!anthropogenic impacts on}  as well as any other kind of maritime works,
e.g. trawling and anchoring of boats \cite{Ribed02}\label{Ribed02}. Other human activities that cause degrading of the
seagrass are recreational boating, commercial overexploitation of coastal resources, eutrophication
\cite{McKenzie06}\label{McKenzie06}.

@@

Besides anthropogenic factors, various biochemical, climatic and environmental processes\index[concepts]{Seagrasses!natural impacts on} can cause
negative influence on seagrass distribution. Seagrass is exposed to threats from the global climate and
environmental change, i.e. increases in sea surface temperature; sea level rise; increased frequency
and intensity of storms and waves; local decrease of water quality, increased sedimentation,
contamination and nitrification; desiccation; salinity fluctuations; nutrient changes; suspended
sediments \cite{Blake00}\label{Blake00}. These stress-drivers can alone result in large-scale seagrass
degradation, but often seagrass undergo simultaneous affects from several of these factors together.
It naturally increases the environmental pressing  to drastic loss of very large areas of
seagrass globally \cite{Orth06}\label{Orth06}.
In tropical areas, where most of seagrasses are located (Fig.\ref{Fig.4}), seagrasses are subjectto catastrophic
extinction and loss, due to the cyclones, typhoons, storms, regular floods and increased rainfalls. Recovery from such
events can take up to several years and often it is only possible by means of the seed reserves from the local
environmental surveys \cite{McKenzie07}\label{McKenzie07}. Other threats for seagrasses in tropical areas are 
increased nutrient availability in the coastal zones, increased
eutrophication and invasive macroalgae. These processes have strong affect on the status of the
seagrass meadows, and often lead to their complete disappearance \cite{Holmer09}\label{Holmer09}.\\
Other environmental threat for the seagrasses arises as a result of the environmental struggle and
competition for existence among species. Thus, meadows of P.oceanica in the Mediterranean Sea are
presently facing invasion by alien algal species, particularly in areas where \textit{P.oceanica} is already
degrading, stressed, have gaps and patchy structure in meadows and show other signs of regression
\cite{Montefalcone10}\label{Montefalcone10}.
Seagrasses are vulnerable fragile species, important for the marine coastal ecosystems, especially for
the protection of the beach structure. However, the facts about seagrass global degrading sound
worrying: about 54 \% of the total seagrass meadows have lost any part of their area; the areas, where
the seagrass ecosystems are degrading or lost, are not located in a specific area or continent, but
registered globally; since 1980s global losses of the seagrasses on our planet is equal to two football
fields per hour \cite{Mellors09b}\label{Mellors09b}.
\pagebreak

@@

\subsection{Research problem}\index[concepts]{Research}
Monitoring\index[concepts]{Research!problem} of the marine benthic ecosystems of seagrasses is essential for the environmental
assessment of the coastal zones. It increases our knowledge of the seagrass ecology, highlights 
threats to the seagrass and preventing them from possible losses and degrading and improves
techniques and methods of the underwater-based observations. Mapping the seagrass contributes to
the evaluating of the seagrass current distribution, analysis of its dynamics and changes over time, as
well as estimations of the degrading of seagrass meadows for the purpose of the coastal management.

\begin{wrapfigure}{r}{0.3\textwidth}
	\centering
	\includegraphics[scale=0.25]{Fig-1-9.jpg}
	\caption{Spectra of the seagrass on different depths (0 - 15 m). Source \cite{Edwards00}\label{Edwards00}}
	\label{fig:9}
\end{wrapfigure}

Precise, correct and up-to-date information about the distribution of \textit{P.oceanica} is necessary for the
sustainable conservation of the marine environment and ecosystems in Mediterranean area, being an
important contribution to the environmental coastal zone management \cite{Pergent-Martini06}\label{Pergent-Martini06}.
However, mapping the seagrass has limitations due to the specific location and characteristics of the
research object. The remote sensing techniques have traditionally been widely used for the seagrass
monitoring. 

@@

Universal, international, standardized methods for seagrass directly for seagrasses
as such still should be developed. Various case studies have been performed, yet their mostly report
methods adjusted for particular areas, without evaluating standard general algorithms that could be
extrapolated towards other regions. \\ Application of the remote sensing towards seagrass mapping is
generally based on the assumption that various types of the seafloor bottom have different
characteristics of the reflectivity, which is visually expressed in distinct colors of the objects. In its
turn, reflectivity of the sediments is affected by the water optical properties and content. For
example, Stephens \cite{Stephens03}\label{Stephens03} prove that microalgae biomass and community structure affect
hyperspectral reflectance of sediments, which enable to estimate total microalgae biomass from
measurements of hyperspectral reflectance. 
Spectral measurements\index[concepts]{Remote sensing!spectral measurements} of the target objects are made
by means of the radiometers, which receive and register the amounts of energy (radiance and
irradiance) from the objects. Measuring optical properties of of the seawater allows to calculate spectra of the
objects and to discriminate them on the aerial and satellite images. Thus, various scientists report
success in spectral discrimination of submerged vegetation and other seafloor cover types on imagery
using hyperspectral optical properties of the sea water for the assessment of benthic habitats \cite{Lewis01,Louchard03,Dehouck08,Werdell03}\label{Lewis01}\label{Louchard03}\label{Dehouck08}\label{Werdell03}.

@@

Studies of spectral reflectance\index[concepts]{Seawater optical properties!spectral reflectance} of the different seagrass species comparing to the spectra of sand and
other seafloor cover types \cite{Vahtmae06}\label{Vahtmae06} prove that spectra of green, brown and red benthic
macroalgae differ from each other, as well as from sand and deep water reflectance spectra. These
differences are well detectable by the means of the remote sensing research methods. Comparing to
the terrestrial plants, aquatic vegetation, \textit{inter alia} seagrass, cannot be detected using red edge of the
spectrum, as these wavelengths are significantly absorbed by water \cite{Kirk94}\label{Kirk94}, as well as by
scattering and absorption by phytoplankton. Some authors \cite{Dierssen03}\label{Dierssen03} also use
spectrally based radiative transfer approach to quantitatively estimate shallow-water bathymetry and
leaf area index (\ac{LAI}) of the seagrass. The spectral reflectance in general is the
result of the spectral absorption in different bands, typical for each target object.
Spectral reflectance\index[concepts]{Seawater optical properties!spectral reflectance} of the seagrass  (Fig.\ref{fig:13}) is largely influenced by the water
depth where it is located, and is generally decreasing in values by increasing depths.
The most important spectral diapason\index[concepts]{Seawater optical properties!spectral reflectance!spectral diapason} for marine mapping of submerged vegetation
and, particularly, for seagrass, is 350-800 nm \cite{Barille09}\label{Barille09}.
Using airborne imagery for retrospective data (before 1970s) together with the most
recent imagery allows to detect changes in seagrass distribution on over different years
and to analyse dynamics of the seagrass distribution \cite{Ball09}\label{Ball09}. \\
Another important advantage of the application of the remote sensing data for mapping of submerged aquatic vegetation has
commercial nature: using remote sensing data and methods enables more low-cost and up-to-date
seagrass mapping \cite{Mumby99}\label{Mumby99}, and is especially useful for the areas where the fieldwork data
capturing is unavailable.
Seagrass meadows may reach spatial scales from several up to hundreds of metres, therefore they are
susceptible by the means of satellite imagery from remote sensors, both with moderate resolution
(e.g., \href{http://landsat.gsfc.nasa.gov/}{Landsat MSS, Landsat TM, Landsat ETM+}, \href{http://envisat.esa.int/instruments/meris/}{MERIS} from \ac{ESA}, \href{http://asterweb.jpl.nasa.gov/}{ASTER}, \href{http://modis.gsfc.nasa.gov/}{MODIS}) and high resolution as
well (e.g., \href{http://www.satimagingcorp.com/gallery-ikonos.html}{IKONOS}, \href{http://www.digitalglobe.com/index.php/85/QuickBird}{Quickbird}, \href{http://www.spotimage.fr/}{SPOT}, \href{http://www.itres.com/products/imagers/casi550}{CASI}). The possibility of their application towards seagrass mapping varies and is limited by the technical characteristics (Table \ref{tab:1}) and resolution of these sensors  (Fig.\ref{fig:2.3}).
In the next paragraphs (\nameref{sec:2.4.1} and \nameref{sec:2.4.2}) we will briefly discuss research experience of the application of satellite imagery towards seagrass mapping, its advantages and limitations, reported by various researchers.
